Modelling BCG vaccination in the UK

### 2019-06-17 rmds/02-background.Rmd
updated background based on conference talk

TB is a global disease with an estimated 10.4 million new cases in 2016, of which 4.3 million were missed by health systems.[@TheWorldHealthOrganization:2018va]  Global incidence rates have decreased year on year since the early 2000's, with an average year on year decrease of 2.9%. However, global TB incidence remains above 134 per 100,000 population (Figure \@ref(fig:global-regional-tb-summary)). On a regional level, incidence rates vary with Africa and South-East Asia having a greater concentration of cases. In the Eastern Mediterranean incidence rates have remained relatively stable over the last 10 years.

@@

Figure \@ref(fig:plot-prop-pul) shows that since the 1980s the proportion of extra-pulmonary TB compared to pulmonary TB has increased from `r prop_extra_pil_1982` in 1982 to `r prop_extra_pil_2016` in 2016. This may be attributed to the age distribution of TB cases changing as different age groups are more likely to progress to pulmonary vs extra-pulmonary TB. It may also be related to the increase of non-UK born cases as a higher proportion of non-UK born cases have extra-pulmonary disease only (51.4%, 2,103/4,089, in 2016), compared to UK born cases (31.9%, 467/1,465, in 2016).[@PHE2017] For more details on TB in England, see Chapter \@ref(data) and the Public Health England 2018 TB report from which the summary data discussed above was extracted.[@PHE2017]

@@

The Bacillus Calmette–Guérin (BCG) vaccine was first given to humans in 1921 and remains the only licensed vaccine for TB.[@Medicine2013] The BCG vaccine is a live vaccine and was developed by weakening a strain of Mycobacterium bovis, which is commonly found in cows, over a period of 13 years.[@TheWorldHealthOrganization:2018va] Serious side effects are rare, although a small scar at the injection site is common. This section details the action, effectiveness, duration of protection, effects and usage of the BCG vaccine.

@@

The effectiveness of the vaccine is impacted by the age at which it is given, the latitude of the individual, and the period of time that has lapsed since vaccination. Multiple randomized control studies (RCTs) have been conducted on BCG efficacy. It has consistently been shown to be highly protective in children.[@Rodrigues1993, @Colditz1994] Efficacy in adults ranges from 0% to 78%,[@Mangtani2014a] with an MRC trial in England finding that BCG was 78% effective.[@Hart1972]

A meta-analysis of RCTs indicated that increased protection is associated with distance from the equator.[@Mangtani2014a] One hypothesis for this is that there is a greater density of mycobacterium near the equator that may mask, or block, the protection offered by the BCG vaccine.[@Zwerling2011] Recently it has been found that much of this latitude effect may be due to stringency in tuberculin skin testing (TST), with lower stringency near the equator.[@TheWorldHealthOrganization:2018va] TST screening tests for the presence of Tuberculosis infection but may give a false positive if the subject has been exposed to other mycobacteria or the BCG vaccine. Reduced stringency would lead to a greater number of TB positive individuals being vaccinated. These individuals would then receive no protection from the vaccine and would lead to a reduced estimate of the effectiveness of the vaccine overall.

@@

Until recently little attention has been given to any additional effects of BCG vaccination.[@Fine2005a; @Teo2006] However, there is now some evidence that BCG vaccination induces innate immune responses that may provide non-specific protection[@Kleinnijenhuis2012] and reduce all-cause neonatal mortality.[@Garly2003; @Higgins] Additionally, BCG vaccination may improve outcomes for individuals with active TB disease. TB patients with BCG scars have been found to respond better to treatment with earlier sputum smear conversion.[@Jeremiah2010] There is also evidence to support an association between BCG vaccination and reduced TB[@Abubakar2013] mortality. The evidence for additional effects of BCG vaccination on outcomes in individuals with notified TB, in England, is explored further in Chapter \@ref(beneficial-bcg-out).

In addition to its effect on TB outcomes, the BCG vaccine has also been found to be effective at preventing leprosy (with an RR of 0.45 (95% CI: 0.34-0.56)), with some evidence that this protection was stronger in those vaccinated before 15 years of age.[@TheWorldHealthOrganization:2018va] Additionally, there is some evidence that the BCG vaccine can provide protection against Non-Tuberculosis mycobacteria infections, with an estimated effectiveness of 50%.[@TheWorldHealthOrganization:2018va]

@@

In England, universal school-aged vaccination was introduced after a MRC trial in the 1950s estimated BCG's effectiveness at 80% in the white UK born population.[@Hart1972] In 2005, the UK shifted from this strategy to targeted vaccination in neonates deemed at high risk.[@Abubakar2013] This change was a reflection of current WHO vaccination policy,[@WHO2017] falling TB incidence rates, an increasing proportion of TB cases occurring in the non-UK born,[@PHE2017] and modelling evidence that suggested stopping the BCG schools scheme would have minimal long term effects on incidence rates.[@Sutherland1989] The impact of this change in policy is explored throughout this thesis but in particular in Chapter \@ref(sutherland), Chapter \@ref(direct-eff) and Chapter \@ref(model-results).

@@

Multiple replacement vaccines are currently in clinical trials.[@Zwerling2011a; @Schrager:2018ip] Vaccine candidates include both live and sub-unit vaccines. Many of these candidate vaccines serve as a boosts to the BCG vaccine, with the BCG vaccine being administered prior to the candidate vaccine.[@Zwerling2011a] Several BCG replacements are also being trialed, both based on alternative methods of attenuating TB mycobacteria and using other approaches.[@Schrager:2018ip] However, in the short-to-midterm it is unlikely that a new vaccine will replace the BCG vaccine. This means that it's optimal usage is as important as ever.

>>>>

TB is a global disease with an estimated 10.4 million new cases in 2016, of which 4.3 million were missed by health systems.[@TheWorldHealthOrganization:2018va]  Global incidence rates have decreased year-on-year since the early 2000's, with an average year on year decrease of 2.9%. However, global TB incidence remains above 134 per 100,000 population (Figure \@ref(fig:global-regional-tb-summary)). On a regional level, incidence rates vary with Africa and South-East Asia having a greater concentration of cases. In the Eastern Mediterranean incidence rates have remained relatively stable over the last 10 years.

@@

Figure \@ref(fig:plot-prop-pul) shows that since the 1980s the proportion of extra-pulmonary TB compared to pulmonary TB has increased from `r prop_extra_pil_1982` in 1982 to `r prop_extra_pil_2016` in 2016. This may be attributed to the age distribution of TB cases changing as different age groups are more likely to progress to pulmonary vs extra-pulmonary TB. It may also be related to the increase of non-UK born cases as a higher proportion of non-UK born cases have extra-pulmonary disease only (51.4%, 2,103/4,089, in 2016), compared to UK born cases (31.9%, 467/1,465, in 2016).[@PHE2017] For more details on TB in England, see Chapter \@ref(data) and the Public Health England 2017 and 2018 TB reports from which the summary data discussed above was extracted.[@PHE2017; @PHE2018]

@@

The Bacillus Calmette–Guérin (BCG) vaccine was first given to humans in 1921 and remains the only licensed vaccine for TB.[@Medicine2013] The BCG vaccine is a live vaccine and was developed by weakening a strain of Mycobacterium bovis, which is commonly found in cows, over a period of 13 years.[@TheWorldHealthOrganization:2018va] Serious side effects are rare, although a small scar at the injection site is common. Initially, public acceptance was slow, with low take up until after the Second World War. However, controversy remains and several countries have scaled back, or redicted, their usage in recent years. This section details the action, effectiveness, duration of protection, effects and usage of the BCG vaccine.

@@

The effectiveness of the vaccine is impacted by the age at which it is given, the latitude of the individual, and the period of time that has lapsed since vaccination. Multiple randomized control studies (RCTs) have been conducted on BCG efficacy. It has consistently been shown to be highly protective in children for both pulmonary TB and TB meningitis.[@Rodrigues1993, @Colditz1994] Efficacy in adults ranges from 0% to 78%,[@Mangtani2014a] with an MRC trial in England finding that BCG was 78% effective in the White UK-born.[@Hart1972] Effectiveness at preventing initial infection has been estimated at 19% (95% CI: 8% to 29%) globally.[@Roy2014]

A meta-analysis of RCTs indicated that increased protection is associated with distance from the equator.[@Mangtani2014a] One hypothesis for this is that there is a greater density of mycobacterium near the equator that may mask, or block, the protection offered by the BCG vaccine.[@Zwerling2011] Recently it has been found that much of this latitude effect may be due to stringency in tuberculin skin testing (TST), with lower stringency near the equator.[@TheWorldHealthOrganization:2018va] TST screening tests for the presence of Tuberculosis infection but may give a false positive if the subject has been exposed to other mycobacteria or the BCG vaccine. Reduced stringency would lead to a greater number of TB positive individuals being vaccinated. These individuals would then receive no protection from the vaccine and would lead to a reduced estimate of the effectiveness of the vaccine overall. Regardless of the mechanism this effect means that vaccination early in life maximises the protection conferred by the BCG vaccine in countries with evidence of decreasing protection with age.

@@

Until recently little attention has been given to any additional effects of BCG vaccination.[@Fine2005a; @Teo2006] However, there is now some evidence that BCG vaccination induces innate immune responses that may provide non-specific protection[@Kleinnijenhuis2012] and reduce all-cause neonatal mortality.[@Garly2003; @Higgins] There is some evidence that this reduction in all-cause mortality extends later in life.[@Rieckmann2016] Additionally, BCG vaccination may improve outcomes for individuals with active TB disease. TB patients with BCG scars have been found to respond better to treatment with earlier sputum smear conversion.[@Jeremiah2010] There is also evidence to support an association between BCG vaccination and reduced TB[@Abubakar2013] mortality. The evidence for additional effects of BCG vaccination on outcomes in individuals with notified TB, in England, is explored further in Chapter \@ref(beneficial-bcg-out).

In addition to its effect on TB outcomes, the BCG vaccine has also been found to be effective at preventing leprosy (with an RR of 0.45 (95% CI: 0.34-0.56)), with some evidence that this protection was stronger in those vaccinated before 15 years of age.[@TheWorldHealthOrganization:2018va] Additionally, there is some evidence that the BCG vaccine can provide protection against Non-Tuberculosis mycobacteria (NTM) infections, with an estimated effectiveness of 50%.[@TheWorldHealthOrganization:2018va] This protection has been shown to greatly reduce the incidence of NTM lymph node inflammation and protect against Buruli ulcers for the first 12 months following vaccination (RR, 0.50 [95% CI, .37–.69]). A study also found that individuals with Buruli ulcers are less likely to develop osteomyelitis (bone infection) if they have a BCG scar (RR, 0.36 (95% CI: 0.22 to 0.58)).[@Zimmermann:2018io]

@@

In England, universal school-aged vaccination was introduced after a MRC trial in the 1950s estimated BCG's effectiveness at 78% in the white UK born population.[@Hart1972] In 2005, the UK shifted from this strategy to targeted vaccination in neonates deemed at high risk.[@Abubakar2013] This change was a reflection of current WHO vaccination policy,[@WHO2017] falling TB incidence rates, an increasing proportion of TB cases occurring in the non-UK born,[@PHE2017] and modelling evidence that suggested stopping the BCG schools scheme would have minimal long term effects on incidence rates.[@Sutherland1989] The impact of this change in policy is explored throughout this thesis but in particular in Chapter \@ref(sutherland), Chapter \@ref(direct-eff) and Chapter \@ref(model-results).

@@

Multiple replacement vaccines are currently in clinical trials.[@Zwerling2011a; @Schrager:2018ip] Vaccine candidates include both live and sub-unit vaccines. Many of these candidate vaccines serve as a boost to the BCG vaccine, with the BCG vaccine being administered prior to the candidate vaccine.[@Zwerling2011a] Several BCG replacements are also being trialed, both based on alternative methods of attenuating TB mycobacteria and using other approaches.[@Schrager:2018ip] However, in the short-to-midterm it is unlikely that a new vaccine will replace the BCG vaccine. This means that it's optimal usage is as important as ever.
